### 2021-04-07 sections/2-info-theory.tex
fixes and typos

The \textbf{distance function} must have certain properties;
specifically, $\forall x, y, z \in X$:
\begin{enumerate}[i.]
    \item $f$ is deterministic.
    \item $f(x, y) = 0 \Leftrightarrow x = y$ and $f(x, y) > 0 \Leftrightarrow x \neq y$.
    \item $f(x, y) = f(y, x)$, i.e. symmetry.
    \item $f(x, z) \leq f(x, y) + f(y, z)$, i.e. the triangle inequality.
\end{enumerate}

@@

We define a \textbf{Metric Ball} as $B(q, r) = \{ x \in X \ | \ f(q, x) \leq r \}$, i.e. a ball containing all instances in the dataset within distance $r$ from an instance $q$.
We also define a \textbf{Cluster} as $C(c, r)$ analogous to a Metric Ball with the caveat that if some clusters have overlapping volumes then any instances in that overlapping volume are assigned to only one of the clusters.
Thus $C_i \cap C_j = \phi \Leftrightarrow i \neq j$ and $\bigcup C_i = X$.

@@

The \textbf{Internal Covering Number} of the dataset, denoted by $N_{\epsilon}^{int}(X)$, is the minimum $n$ such that $\{x_1, \dots, x_n\} \subset X$ and $\bigcup_{i = 1}^{n} C(c_i, \epsilon) = X$, i.e. the clusters $C_i$ collectively cover $X$.

Given an \textbf{Internal Covering} of $X$, we can specify any $\bar{x} \in X$ to within error $\epsilon$ by specifying $x_i$ such that $\bar{x} \in B(x_i, \epsilon)$.
This it takes $\mathcal{O} \big( \log N_{\epsilon}^{ent}(X) \big)$ bits of information to specify $\bar{x}$ to within error $\epsilon$.

Given two instances $x_1, x_2 \in X$, we can construct $x_2 - x_1$, a \textbf{minimal encoding} of $x_2$ in terms of $x_1$, such that the number of bits needed to specify the encoding scales linearly with $f(x_1, x_2)$.
We would need to construct a method for construction such \textit{minimal encodings} for each pairing of $(X, f)$.
See Sections~\ref{subsec:methods:compression} and~\ref{sec:datasets-and-distance-functions} for some examples.

The \textbf{Shannon Entropy} of an instance $x$, denoted by $H(x)$, is the number of bits needed to exactly specify $x$.
Given an encoding, $x_2 - x_1$, of an instance $x_2$ in terms of another instance $x_1$, let $H(x_2 - x_1)$ be the Shannon Entropy of the encoding, i.e. the number of bits needed to store the encoding.
Therefore, \[ H \big( \{ x_1, x_2 \} \big) \leq H(x_1) + H(x_2 - x_1) \] where $\{ x_1, x_2 \}$ is the concatenation of $x_1$ and $x_2$.

For example, given a dataset of genomic strings with the Levenshtein edit distance as the metric, let $x_1$ and $x_2$ be two strings in the dataset.
Let $x_2 - x_1$ be the shortest list of edits to convert $x_1$ into $x_2$.
The length of this edit list is equal to the Levenshtein edit distance i.e. $|x_2 - x_1| = f(x_1, x_2)$.
As in~\cite{berger2020levenshtein}, each such edit can be stored using $\mathcal{O}(\log m)$ bits where $m$ is the maximum length of any string in the dataset.
Thus, the total number of bits needed for each such encoding is $\mathcal{O} \big( f(x_1, x_2) \big)$.

Hence, for $|X| < \infty$,
\begin{align*}
    H (C) &\leq H(c) + \sum_{x \in C} H(x - c) \\
    &\leq H(c) + |C| \cdot H(y - c) \\ \\
    \text{where} \ y &= \argmax_{x \in C} \ f(x, c) \text{, or equivalently} \ f(y, c) = r.
\end{align*}

For a \textbf{Hierarchical Clustering}, let the cluster $C$ have parent $C_{parent}$, children $C_{left}$ and $C_{right}$, center $c \in C$, radius $r$, and an instance $y \in C$ such that $f(y, c) = r$.

We have two choices for encoding the instances in a cluster:
either each instance can be encoded in terms of the cluster center, 
or each child-cluster can be encoded in terms of the child-center and the child-centers encoded in terms of the cluster-center.

@@

Let $2r$ be the maximum pairwise distance among the instances in the dataset.
In the worst case, e.g. with the distribution shaped like a $d$-sphere, our chosen axes will be such that $2r$ is the maximum pairwise distance \textit{along every axis}.
Such a distribution would produce a balanced clustering.

Partition will select a maximally distant pair of points to use as poles, i.e. it will choose one of the $d$ axes along which to split the cluster into two children.
After one Partition, the maximum pairwise distance along that axis will be bounded above by $r$.
The next recursive Partition will select another of the $d$ axes.
Thus, after at most $d$ Partitions, the maximum pairwise distance along each axis will be bounded above by $r$.
The overall maximum pairwise distance, \textit{i.e. not restricted to be along one axis}, will be bounded above by $r \sqrt{2}$, for example, between two instances that lie at the extrema of different axes.

@@

Note that, in practice, we never see a balanced clustering.
Instead, Partition produces unbalanced trees due the varying density of sampling in different regions of the manifold.
Further, the cluster radii decrease by a factor much larger than $\frac{1}{\sqrt{2}}$ and they do so every one or two partitions rather than after $d$ partitions.
See Section~\ref{sec:results} for results on several datasets.

@@

In the general case, let $T(C, d)$ be the subtree rooted at $C$ with depth $d$ and up-to $2^d$ leaves.
Since we cannot partition a cluster containing only one instance, we can bound the number of leaves, i.e. $2^d \leq |C|$.
Let the leaves be indexed with integers $j \in [1, 2^{d + 1} - 1]$ as obtained by a breadth-first traversal of the tree, i.e. the root has index $1$ and a cluster with index $i$ has children with indices $2i$ and $2i + 1$.
Let the center of each child-cluster be encoded in terms of the center of its parent-cluster.
Therefore, $H(T)$, the Shannon Entropy of $T(C, d)$, is given by
\begin{align*}
    H(T) &= H(c_2 - c_1) + H(c_3 - c_1) + H(c_4 - c_2) + H(c_5 - c_2) \\
    & \ \ \ \ + H(c_6 - c_3) + H(c_7 - c_3) + \dots + H(c_{2^{d + 1} - 1} - c_{2^d})
\end{align*}
i.e. the sum of all encodings of child-centers in terms of their parent-centers, where $c_1$ is the center of the root cluster.

@@

We can finally rewrite Equation~\ref{eq:hierarchical-shannon-entropy} as
\begin{align}
    \label{eq:hierarchical-shannon-entropy-2}
    H(C) &\leq H(c - c_{parent}) + \min \begin{cases}
        |C| \cdot H(y - c) \\
        |C| \cdot H(y - c) \cdot \frac{1}{\sqrt{2}} + H(T)
    \end{cases}
\end{align}

@@

When starting out encoding the dataset in this fashion, for low depths in the tree, we expect that using subtrees to encode the instances will lead to great savings in the number of bits used.
As we continue deeper into the tree, we would start seeing diminishing returns for using subtrees.
Eventually, either the difference will be so small that it is not worth the computational cost of Partition, or the two cases will equalize, or the first case will become the better choice.
At such a depth in the tree, we can encode the instances in a cluster using the cluster-center rather than the subtree.
That cluster-center can be encoded in terms of the center of its parent and so on back up the tree to the root.

>>>>

The \textbf{distance function} must have must be such that $\forall x, y, z \in X$:
\begin{enumerate}[i.]
    \item $f$ is deterministic.
    \item $f(x, y) = 0 \Leftrightarrow x = y$ and $f(x, y) > 0 \Leftrightarrow x \neq y$.
    \item $f(x, y) = f(y, x)$, i.e. symmetry.
    \item $f(x, z) \leq f(x, y) + f(y, z)$, i.e. the triangle inequality.
\end{enumerate}

@@

We define a \textbf{Metric Ball} as $B(q, r) = \{ x \in X \ | \ f(q, x) \leq r \}$, i.e. a ball containing all instances in the dataset within distance $r$ from an instance $q$.
We also define a \textbf{Cluster} as $C(c, r)$ analogous to a Metric Ball with the caveat that if some clusters have overlapping volumes then any instances in that overlapping volume are assigned to only one of the clusters.
Thus $\bigcup C_i = X$ and $i \neq j \Rightarrow C_i \cap C_j = \phi$.

The \textbf{Internal Covering Number} of the dataset, denoted by $N_{\epsilon}^{int}(X)$, is the minimum $n$ such that $\{x_1, \dots, x_n\} \subset X$ and $\bigcup_{i = 1}^{n} C(c_i, \epsilon) = X$, i.e. the minimum number of $\epsilon$-radius clusters that collectively cover $X$.

Given an \textbf{Internal Covering} of $X$, we can specify any $\bar{x} \in X$ to within error $\epsilon$ by specifying $x_i$ such that $\bar{x} \in B(x_i, \epsilon)$.
Thus it takes $\mathcal{O} \big( \log N_{\epsilon}^{ent}(X) \big)$ bits of information to specify $\bar{x}$ to within error $\epsilon$.

Given two instances $x_1, x_2 \in X$, we can construct $x_2 - x_1$, a \textbf{minimal encoding} of $x_2$ in terms of $x_1$, such that the number of bits needed to specify the encoding scales linearly with $f(x_1, x_2)$.
We would need a potentially different method for constructing such \textit{minimal encodings} for each pairing of $(X, f)$.
See Sections~\ref{subsec:methods:compression} and~\ref{sec:datasets-and-distance-functions} for some examples.

@@

The \textbf{Shannon Entropy} of an instance $x$, denoted by $H(x)$, is the number of bits needed to exactly specify $x$.

Given an encoding, $x_2 - x_1$, of an instance $x_2$ in terms of another instance $x_1$, let $H(x_2 - x_1)$ be the Shannon Entropy of the encoding, i.e. the number of bits needed to store the encoding.
Therefore,
\begin{align*}
    H \big( \{ x_1, x_2 \} \big) \leq H(x_1) + H(x_2 - x_1)
\end{align*}
where $\{ x_1, x_2 \}$ is the concatenation of $x_1$ and $x_2$.

For example, given a dataset of genomic strings with the Levenshtein edit distance as the metric, let $x_1$ and $x_2$ be two strings in the dataset.
Let $x_2 - x_1$ be a (possibly non-unique) shortest list of edits to convert $x_1$ into $x_2$.
The length of this edit list is equal to the Levenshtein edit distance i.e. $|x_2 - x_1| = f(x_1, x_2)$.
As in~\cite{berger2020levenshtein}, each such edit can be stored using $\mathcal{O}(\log m)$ bits where $m$ is the maximum length of any string in the dataset.
Thus, the total number of bits needed for each such encoding is $\mathcal{O} \big( f(x_1, x_2) \big)$.

Let $H(C)$ be the Shannon Entropy of the cluster $C(c, r)$.
Hence,
\begin{align*}
    H (C) &\leq H(c) + \sum_{x \in C} H(x - c) \\
    &\leq H(c) + |C| \cdot H(y - c) \\ \\
    \text{where} \ y &= \argmax_{x \in C} \ f(x, c) \text{, or equivalently} \ f(y, c) = r.
\end{align*}

For a \textbf{Hierarchical Clustering}, let the cluster $C$ have 
parent $C_{parent}$, 
children $C_{left}$ and $C_{right}$, 
center $c \in C$, 
radius $r \in \mathbb{R}^+$, 
and an instance $y \in C$ such that $f(y, c) = r$.

We have two choices for encoding the instances in a cluster:
either each instance can be encoded in terms of the cluster-center, 
or each child-cluster can be encoded in terms of the child-center and the child-centers encoded in terms of the cluster-center.

@@

Let $2r$ be the maximum pairwise distance among the instances in the dataset.
We choose the axes such that the two points that are $2r$ apart lie along one of the axes.
Thus, a $d$-dimensional hyper-sphere of radius $r$ would bound the dataset.
In the worst case, e.g. with a distribution that fills the $d$-sphere, our axes will be such that $2r$ is the maximum pairwise distance \textit{along every axis}.
Such a distribution would also produce a balanced clustering.

Partition will select a maximally distant pair of points to use as poles, i.e. it will choose one of the $d$ axes along which to split the cluster into two children.
After one Partition, the maximum pairwise distance along that axis will be bounded above by $r$.
The next recursive Partition will select another of the $d$ axes.
Thus, after at most $d$ Partitions, the maximum pairwise distance along each axis will be bounded above by $r$.
The overall maximum pairwise distance, i.e. \textit{not restricted} to be along one axis, will be bounded above by $r \sqrt{2}$ by, for example, two instances that lie at the extrema of different axes.

@@

Note that, in practice, we never see a balanced clustering.
Instead, Partition produces unbalanced trees due to the varying density of sampling in different regions of the manifold and the low-dimensional ``shape'' of the manifold.
Further, the cluster radii decrease by a factor much larger than $\frac{1}{\sqrt{2}}$ and they do so every one or two partitions rather than after $d$ partitions.
See Section~\ref{sec:results} for observations of radii-scaling on several datasets.

@@

In the general case, let $T(C, d)$ be the subtree rooted at $C$ with depth $d$ and up-to $2^d$ leaves.
Since we cannot partition a cluster containing only one instance, we can bound the number of leaves, i.e. $2^d \leq |C|$.
Let the leaves be indexed with integers $j \in [1, 2^{d + 1} - 1]$ as obtained by a breadth-first traversal of the tree, i.e. the root has index $1$, a cluster with index $i < 2^d$ has children with indices $2i$ and $2i + 1$, and a cluster with index $i \geq 2^d$ is a leaf-cluster with no children.
Let the center of each child-cluster be encoded in terms of the center of its parent-cluster.
Therefore, $H(T)$, the Shannon Entropy of $T(C, d)$, is given by
\begin{align*}
    H(T) &= H(c_2 - c_1) + H(c_3 - c_1) + H(c_4 - c_2) + H(c_5 - c_2) \\
    & \ \ \ \ + H(c_6 - c_3) + H(c_7 - c_3) + \dots + H(c_{2^{d + 1} - 1} - c_{2^d})
\end{align*}
i.e. the sum of all encodings of child-centers in terms of their parent-centers, where $c_1$ is the center of the root cluster.

@@

Finally, we rewrite Equation~\ref{eq:hierarchical-shannon-entropy} as
\begin{align}
    \label{eq:hierarchical-shannon-entropy-2}
    H(C) &\leq H(c - c_{parent}) + \min \begin{cases}
        |C| \cdot H(y - c) \\
        |C| \cdot H(y - c) \cdot \frac{1}{\sqrt{2}} + H(T)
    \end{cases}
\end{align}

@@

When starting out encoding the dataset in this fashion, for low depths in the tree, we expect that using subtrees to encode the instances will lead to great savings in the number of bits used.
As we continue deeper into the tree, we would start seeing diminishing returns for using subtrees.
Eventually, either 
the difference will be so small so as not to be worth the computational cost of Partition, 
or the two cases will equalize, 
or the first case will become the better choice.
At such a depth in the tree, we can encode the instances in a cluster using the cluster-center rather than the subtree.
The center of that cluster can be encoded in terms of the center of its parent and so on back up the tree to the root.
